Rectangulotopes

### 2024-10-14 draft/rectangulotopes.tex
some corrections from report Y

\begin{figure}
	\capstart
	\centerline{\includegraphics[scale=.75]{permutahedronLeft4} \quad \raisebox{-.5cm}{\includegraphics[scale=.75]{associahedron4}}}
	\caption{The permutahedron~$\Perm(4)$ (left) and the associahedron~$\Asso(4)$ (right). \cite[Fig.~1]{MR3964495} \&~\cite[Fig.~5]{MR4584712}}
	\label{fig:quotientopes}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\textwidth]{rectangulations.pdf}
	\caption{Four rectangulations. The rightmost rectangulation is not weakly equivalent to any other. The leftmost three are weakly equivalent, but only the first two are strongly equivalent. Only the third one is diagonal.}
	\label{fig:rectequiv}
\end{figure}

@@

Weak and strong rectangulations of size $n$ have been shown to define congruences of the weak Bruhat order on $\f{S}_n$.

@@

Recall that the \defn{inorder labeling} of a binary tree~$T$ is the labeling of the nodes of~$T$ such that the label of each node~$t$ of~$T$ is larger than all labels in the left subtree of~$t$ and larger than all labels in the right subtree of~$t$.
For any rectangle~$r$ of~$R$, the inorder label of the source~$s(r)$ in~$S(R)$ coincides with the inorder label of the target~$t(r)$ in~$T(R)$.
The inorder can also be retrieved from the above--below and left--right relationships defined before, as the transitive closure of the union of the below and right partial orders.
This enables to unambiguously label the rectangles of~$R$ by the inorder.
The resulting labeling coincides with the NW--SE labeling of~\cite{ACFF24}.
From now on, the labels of the rectangles in a rectangulation are the inorder labels, and the two trees~$T(R)$ and $S(R)$ are defined on the vertex set $[n]$ accordingly.
For each $i\in [n]$, we refer to its two subtrees in $T(R)$ as the \defn{horizontal} and \defn{vertical subtrees}, and similarly for $S(R)$.

@@

In order to exhibit a similar realization for the strong rectangulotopes, we need to count subsets of leaves that lie in some subtrees of both the source and target trees of a rectangulation.
%
Two leaves of the trees $T(R)$ and $S(R)$ are said to be \defn{common leaves} if the edges to their parents lie on the same segment of $R$.
%
We denote by ${}^i \!\!\! \raisebox{.03cm}{\verticalPattern} \!\!\! {}_j$ the situation where the rectangles $r_i$ and $r_j$ of inorder labels $i$ and $j$ share a vertical segment containing the right edge of~$r_i$ and the left edge of~$r_j$, and the bottom edge of $r_i$ is strictly below the top edge of~$r_j$.
Similarly, we denote by~$\; {}^i \!\!\!\!\! \horizontalPattern \!\!\!\!\! \raisebox{-.06cm}{$_j$} \,$ the situation where the rectangles $r_i$ and $r_j$ share a horizontal segment containing the bottom edge of~$r_i$ and the top edge of~$r_j$, and the right edge of~$r_i$ is strictly on the right of the left edge of~$r_j$.
We use the Iverson bracket $\llbracket \varphi\rrbracket$, which is equal to 1 if $\varphi$ holds, and 0 otherwise.

\begin{theorem}
\label{thm:strongRectangulotope}
% The strong rectangulotope is realized by the polytope~$\SRP (n)$ defined equivalently as
The strong rectangulotope~$\SRP (n)$ is realized by the polytope equivalently described as:
  \begin{enumerate}[(i)]
  \item the convex hull of the points
  \[
  \sum_{i,j\in [n], i< j} (\yin{w}^R_{i,j} - \yang{w}^R_{i,j})\cdot (\b{e}_i - \b{e}_j),
  \]
   for all strong rectangulations $R$ of size~$n$, with
  \[
    \yin{w}^R_{i,j} \eqdef h^T_i \cdot cv^{T,S}_{i,j}\cdot h^S_ j\cdot \llbracket \; \neg \; {}^i \!\!\! \raisebox{.03cm}{\verticalPattern} \!\!\! {}_j \; \rrbracket
    \qquad\text{and}\qquad
    \yang{w}^R_{i,j} \eqdef v^S_i \cdot ch^{S,T}_{i,j}\cdot v^T_j \cdot \llbracket \; \neg \;\; {}^i \!\!\!\!\! \horizontalPattern \!\!\!\!\! \raisebox{-.06cm}{$_j$} \;\; \rrbracket,
  \]
 where:
  \begin{itemize}
  \item $S$ and $T$ are the source and target trees of the rectangulation~$R$,
  \item $h^T_i$ and $v^T_i$ denote the number of leaves in the horizontal and vertical subtrees of $i$ in~$T$,
  \item $ch^{T,S}_{i,j}$ (resp.~$cv^{T,S}_{i,j}$) denote the number of common leaves of the horizontal (resp.~vertical) subtree of $i$ in $T$ and the horizontal (resp.~vertical) subtree of $j$ in $S$,
  \end{itemize}
\item the intersection of the hyperplane defined by~$\sum_{i \in [n]} x_i = 0$ with the halfspaces defined by
  \[
  \sum_{i \in X} x_i \le \#\set{I,J}{I \not\subseteq X \text{ and } J \not\subseteq [n] \ssm X} + \#\set{I,J}{I \not\subseteq [n] \ssm X \text{ and } J \not\subseteq X}
  \]
  for all subsets~$\varnothing \ne X \subsetneq [n]$, where~$I,J$ denote nonempty and consecutive intervals of~$[n]$.
  \end{enumerate}
\end{theorem}

@@

\enlargethispage{.2cm}
We think of the formula of \cref{thm:strongRectangulotope}\,(i) as the analogue of J.-L.-Loday's product formula of \cref{thm:loday}\,(i) for the coordinates of the vertices of the associahedron.
We note however that our formula is quadratic in~$n$.
It remains unclear to us if there is a simple linear formula to compute the coordinates of the vertex of~$\SRP(n)$ corresponding to a given strong rectangulation.

@@

We now briefly recall some results on the weak Bruhat order and its quotients, both from a lattice and geometric perspectives.
We refer to \cite{MR2142177, Reading-arcDiagrams, MR3645056, MR3645055, MR3964495, MR4584712} for details.

@@

\pagebreak 
A set function $f:2^{[n]}\mapsto \R$ is said to be \defn{submodular} when $f(X \cup Y) + f(X \cap Y) \le f(X) + f(Y)$ for any~$X, Y \subseteq [n]$.
By definition, any quotientope is a deformed permutahedron, and can thus be written as
  \[
  \Bigset{x \in \R^n}{ \sum_{i \in [n]} x_i = f([n]) \text{ and } \sum_{i \in X} x_i \le f(X) \text{ for all } \varnothing \ne X \subseteq [n]},
  \]
where~$f$ is a submodular set function.
We will need the following simple result: Given a collection $\polytope{P}_1, \polytope{P}_2,\ldots ,\polytope{P}_k$ of deformed permutahedra defined by submodular set functions $f_1,f_2,\ldots ,f_k$, their Minkowski sum $\sum_i \polytope{P}_i$ is defined by the submodular function $f \eqdef \sum_i f_i$.

@@

\subsection{The weak rectangulation congruence}
\label{subsec:weakRectangulationCongruence}

@@

\pagebreak
\begin{theorem}[\cite{MR2864445,ACFF24}]
  \label{thm:strongCong}
  The subsets $\mathcal{L}(\prec_s^R)$ of $\f{S}_n$, for all rectangulations $R$ of size~$n$, are equivalence classes of a congruence relation $\strongeq$ on the weak Bruhat order on $\f{S}_n$.
    Furthermore:
  \begin{itemize}
  \item the congruence classes are one-to-one with strong rectangulations,  
  \item the minimal elements of the congruence classes are the 2-clumped permutations,
  \item the maximal elements of the congruence classes are the co-2-clumped permutations.
  \end{itemize}
\end{theorem}

@@

\begin{lemma}
  The strong rectangulotope $\SRP(n)$ is realized by the following submodular function:
  \[
  f_{\strongeq}(X) \eqdef \# \bigset{ I, J }{ I \not\subseteq [n] \ssm X \text{ and } J \not\subseteq X } +
  \#\bigset{ I, J }{ I \not\subseteq X \text{ and } J \not\subseteq [n] \ssm X },
  \]
  where $I, J$ are pairs of nonempty consecutive intervals of~$[n]$.
\end{lemma}

@@

\enlargethispage{.3cm}
Note again that by definition, this extremal vertex should only depend on the strong congruence class of $\sigma$.
This, again, gives us alternative characterizations of the yin congruence $\sim$, the yang congruence $\backsim$, and the strong rectangulation congruence $\strongeq$.
We have $\sigma \sim \tau$ if and only if for any pair of consecutive intervals $I$ and $J$ of $[n]$, 
$\arg\max_{k\in I} \sigma^{-1}_k = \arg\max_{k\in I} \tau^{-1}_k$ and 
$\arg\min_{k\in J} \sigma^{-1}_k = \arg\min_{k\in J} \tau^{-1}_k$ (that is, if $\sigma \weakeq \tau$),
and, if we let $i$ and $j$ be defined as in Lemma~\ref{lem:yinminmax} as the indices of the maximum in $I$ and the minimum in $J$,
$\sigma^{-1}_i > \sigma^{-1}_j \Leftrightarrow \tau^{-1}_i > \tau^{-1}_j$.
A similar statement holds for the yang congruence $\backsim$, by exchanging $\min$ and $\max$.
Finally, $\sigma\strongeq\tau$ if both sets of conditions hold.

>>>>

\begin{figure}
	\capstart
	\centerline{\includegraphics[scale=.75]{permutahedronLeft4} \quad \raisebox{-.5cm}{\includegraphics[scale=.75]{associahedron4}}}
	\caption{The permutahedron~$\Perm(4)$ (left) and the associahedron~$\Asso(4)$ (right). \cite[Fig.~1]{MR3964495} \&~\cite[Fig.~5]{MR4584712}}
	\vspace{-.4cm}
	\label{fig:quotientopes}
\end{figure}

@@

\begin{figure}
  \includegraphics[width=\textwidth]{rectangulations.pdf}
	\caption{Four rectangulations. The rightmost rectangulation is not weakly equivalent to any other. The leftmost three are weakly equivalent, but only the first two are strongly equivalent. Only the third one is diagonal.}
	\vspace{-.5cm}
	\label{fig:rectequiv}
\end{figure}

@@

\enlargethispage{.2cm}
Weak and strong rectangulations of size $n$ have been shown to define congruences of the weak Bruhat order on $\f{S}_n$.

@@

Recall that the \defn{inorder labeling} of a binary tree~$T$ is the labeling of the nodes of~$T$ such that the label of each node~$t$ of~$T$ is larger than all labels in the left subtree of~$t$ and smaller than all labels in the right subtree of~$t$.
For any rectangle~$r$ of~$R$, the inorder label of the source~$s(r)$ in~$S(R)$ coincides with the inorder label of the target~$t(r)$ in~$T(R)$.
The inorder can also be retrieved from the above--below and left--right relationships defined before, as the transitive closure of the union of the below and right partial orders.
This enables to unambiguously label the rectangles of~$R$ by the inorder.
The resulting labeling coincides with the NW--SE labeling of~\cite{ACFF24}.
From now on, the labels of the rectangles in a rectangulation are the inorder labels, and the two trees~$T(R)$ and $S(R)$ are defined on the vertex set $[n]$ accordingly.
For each vertex in $T(R)$ (or in~$S(R)$), we refer to its subtree reached by a horizontal (resp.~vertical) edge as its \defn{horizontal} (resp.~\defn{vertical})~\defn{subtree}.

@@

In order to exhibit a similar realization for the strong rectangulotopes, we need to count subsets of leaves that lie in some subtrees of both the source and target trees of a rectangulation.
%
Two leaves of the trees $T(R)$ and $S(R)$ are said to be \defn{common leaves} if the edges to their parents lie on the same segment of $R$.
%
We denote by ${}^i \!\!\! \raisebox{.03cm}{\verticalPattern} \!\!\! {}_j$ the situation where the rectangles $r_i$ and $r_j$ of inorder labels $i$ and $j$ share a vertical segment containing the right edge of~$r_i$ and the left edge of~$r_j$, and the bottom edge of $r_i$ is strictly below the top edge of~$r_j$.
Similarly, we denote by~$\; {}^i \!\!\!\!\! \horizontalPattern \!\!\!\!\! \raisebox{-.06cm}{$_j$} \,$ the situation where the rectangles $r_i$ and $r_j$ share a horizontal segment containing the bottom edge of~$r_i$ and the top edge of~$r_j$, and the right edge of~$r_i$ is strictly on the right of the left edge of~$r_j$.
We use the Iverson bracket $\llbracket \varphi\rrbracket$, which is equal to 1 if $\varphi$ holds, and 0 otherwise, and~$\neg \, \varphi$ for the logical negation of~$\varphi$.

\begin{theorem}
\label{thm:strongRectangulotope}
% The strong rectangulotope is realized by the polytope~$\SRP (n)$ defined equivalently as
The strong rectangulotope~$\SRP (n)$ is realized by the polytope equivalently described as:
  \begin{enumerate}[(i)]
  \item the convex hull of the points
  \[
  \sum_{i,j\in [n], i< j} (\yin{w}^R_{i,j} - \yang{w}^R_{i,j})\cdot (\b{e}_i - \b{e}_j),
  \]
   for all strong rectangulations $R$ of size~$n$, with
  \[
    \yin{w}^R_{i,j} \eqdef h^T_i \cdot cv^{T,S}_{i,j}\cdot h^S_ j\cdot \llbracket \; \neg \; {}^i \!\!\! \raisebox{.03cm}{\verticalPattern} \!\!\! {}_j \; \rrbracket
    \qquad\text{and}\qquad
    \yang{w}^R_{i,j} \eqdef v^S_i \cdot ch^{S,T}_{i,j}\cdot v^T_j \cdot \llbracket \; \neg \;\; {}^i \!\!\!\!\! \horizontalPattern \!\!\!\!\! \raisebox{-.06cm}{$_j$} \;\; \rrbracket,
  \]
 where:
  \begin{itemize}
  \item $S$ and $T$ are the source and target trees of the rectangulation~$R$,
  \item $h^T_i$ and $v^T_i$ denote the number of leaves in the horizontal and vertical subtrees of $i$ in~$T$,
  \item $ch^{T,S}_{i,j}$ (resp.~$cv^{T,S}_{i,j}$) denote the number of common leaves of the horizontal (resp.~vertical) subtree of $i$ in $T$ and the horizontal (resp.~vertical) subtree of $j$ in $S$,
  \end{itemize}
\item the intersection of the hyperplane defined by~$\sum_{i \in [n]} x_i = 0$ with the halfspaces defined by
  \[
  \sum_{i \in X} x_i \le \#\set{I,J}{I \not\subseteq X \text{ and } J \not\subseteq [n] \ssm X} + \#\set{I,J}{I \not\subseteq [n] \ssm X \text{ and } J \not\subseteq X}
  \]
  for all subsets~$\varnothing \ne X \subsetneq [n]$, where~$I,J$ are nonempty consecutive intervals of~$[n]$, meaning that~$\max(I) = \min(J)-1$.
  \end{enumerate}
\end{theorem}

@@

We think of the formula of \cref{thm:strongRectangulotope}\,(i) as the analogue of J.-L.-Loday's product formula of \cref{thm:loday}\,(i) for the coordinates of the vertices of the associahedron.
We note however that our formula is quadratic in~$n$.
It remains unclear to us if there is a simple linear formula to compute the coordinates of the vertex of~$\SRP(n)$ corresponding to a given strong rectangulation.

@@

We now briefly recall some results on the weak Bruhat order and its quotients, both from a lattice and geometric perspective.
We refer to \cite{MR2142177, Reading-arcDiagrams, MR3645056, MR3645055, MR3964495, MR4584712} for details.

@@

A set function $f:2^{[n]}\mapsto \R$ is said to be \defn{submodular} when $f(X \cup Y) + f(X \cap Y) \le f(X) + f(Y)$ for any~$X, Y \subseteq [n]$.
By definition, any quotientope is a deformed permutahedron, and can thus be written as
  \[
  \Bigset{x \in \R^n}{ \sum_{i \in [n]} x_i = f([n]) \text{ and } \sum_{i \in X} x_i \le f(X) \text{ for all } \varnothing \ne X \subseteq [n]},
  \]
where~$f$ is a submodular set function.
We will need the following simple result: Given a collection $\polytope{P}_1, \polytope{P}_2,\ldots ,\polytope{P}_k$ of deformed permutahedra defined by submodular set functions $f_1,f_2,\ldots ,f_k$, their Minkowski sum $\sum_i \polytope{P}_i$ is defined by the submodular function $f \eqdef \sum_i f_i$.

@@

\pagebreak
\subsection{The weak rectangulation congruence}
\label{subsec:weakRectangulationCongruence}

@@

\begin{theorem}[\cite{MR2864445,ACFF24}]
  \label{thm:strongCong}
  The subsets $\mathcal{L}(\prec_s^R)$ of $\f{S}_n$, for all rectangulations $R$ of size~$n$, are equivalence classes of a congruence relation $\strongeq$ on the weak Bruhat order on $\f{S}_n$.
    Furthermore:
  \begin{itemize}
  \item the congruence classes are one-to-one with strong rectangulations,  
  \item the minimal elements of the congruence classes are the 2-clumped permutations,
  \item the maximal elements of the congruence classes are the co-2-clumped permutations.
  \end{itemize}
\end{theorem}

@@

\begin{lemma}
  The strong rectangulotope $\SRP(n)$ is realized by the following submodular function:
  \[
  f_{\strongeq}(X) \eqdef \# \bigset{ I, J }{ I \not\subseteq [n] \ssm X \text{ and } J \not\subseteq X } +
  \#\bigset{ I, J }{ I \not\subseteq X \text{ and } J \not\subseteq [n] \ssm X },
  \]
  where $I, J$ are nonempty consecutive intervals of~$[n]$.
\end{lemma}

@@

Note again that by definition, this extremal vertex should only depend on the strong congruence class of $\sigma$.
This, again, gives us alternative characterizations of the yin congruence $\sim$, the yang congruence $\backsim$, and the strong rectangulation congruence $\strongeq$.
We have $\sigma \sim \tau$ if and only if for any two consecutive intervals $I$ and $J$ of $[n]$, 
$\arg\max_{k\in I} \sigma^{-1}_k = \arg\max_{k\in I} \tau^{-1}_k$ and 
$\arg\min_{k\in J} \sigma^{-1}_k = \arg\min_{k\in J} \tau^{-1}_k$ (that is, if $\sigma \weakeq \tau$),
and, if we let $i$ and $j$ be defined as in Lemma~\ref{lem:yinminmax} as the indices of the maximum in $I$ and the minimum in $J$,
$\sigma^{-1}_i > \sigma^{-1}_j \Leftrightarrow \tau^{-1}_i > \tau^{-1}_j$.
A similar statement holds for the yang congruence $\backsim$, by exchanging $\min$ and $\max$.
Finally, $\sigma\strongeq\tau$ if both sets of conditions hold.
